Impact of grids and dynamical cores in CESM2.2 on the surface mass balance of the Greenland Ice Sheet

### 2022-04-06 arcticgrids.tex
Updates from Overleaf

Before delving into the simulated Arctic, we assess the global mean differences between the various grids and dycores. Figure~\ref{fig:dT-lores} shows 1979-1998 annual mean, zonal mean height plots expressed as differences between uniform-resolution grids and dycores. The $f09$ grid is warmer than the $f19$ grid, primarily in the mid-to-high latitudes throughout the depth of the troposphere. This is a common response to increasing horizontal resolution in GCMs \cite{PS2002CD,RETAL2006JC}.
\citeA{HK2020QJRMS} have shown that this occurs in CAM due to greater resolved vertical velocities which, in turn, drive more condensational heating in the CLUBB macrophyiscs. The right columns in Figure~\ref{fig:dT-lores} support this interpretation, showing an increase in the climatological CLUBB heating in low and mid-latitudes on the $f09$ grid.

@@

It is useful to understand summer temperature biases due to their control on ice and snow melt over the Greenland Ice Sheet \cite<GrIS;>{O2001JAM}. Figure~\ref{fig:dThyps} shows the 1979--1998 lower troposphere summer temperature bias relative to ERA5. It is computed from the 500--1000 hPa geopotential thickness, solving for the layer-mean virtual temperature using the hypsometric equation. The results are consistent with the zonal mean height plots; increasing resolution from $f19$ to $f09$ warms the climate, and the $1^{\circ}$ SE grids are warmer than the FV grids. The FV summer temperatures are persistently colder than ERA5, whereas the $1^{\circ}$ SE grids are not as cold, and are actually warmer than ERA5 at high-latitudes, north of $75^{\circ}$. All grids show a north-south gradient in bias over Greenland, with the summer temperature bias more positive for the northern part of the ice sheet. This pattern is also evident in the 2m summer temperature bias over Greenland (not shown).

@@

Some of these temperature differences may be related to different summer shortwave cloud forcing for the various grids and dycores. Figure~\ref{fig:SWCF} shows the summer shortwave cloud forcing bias in six runs, using the CERES-EBAF product. 
A negative bias corresponds to excessive reflection and cooling.
The uniform grids have similar biases, with the clouds reflecting 20--40 W/m$^2$ too much shortwave radiation over a wide swath of the Arctic, primarily the land masses. There is also a halo of positive bias (clouds not reflective enough) around the ocean perimeter of Greenland. The $Arctic$ grid has much smaller cloud forcing biases over the Arctic land masses, but is still too reflective over Alaska, the Canadian Archipelago, and parts of Eurasia.  Compared to the $Arctic$ grid, the $ARCTCGRIS$ grid vastly reduces the cloud forcing bias over Eurasia, and also improves the bias over North America. In both VR grids, the halo of positive forcing bias around the perimeter of Greenland is absent.

While the summer cloud forcing biases are consistent with the summer temperature biases in Figure~\ref{fig:dThyps} -- regions where clouds are too reflective coincide with regions that are too cold -- it is not clear whether the cold biases are caused by the cloud biases, or whether the cold biases amplify the cloud forcing bias.

@@

The accuracy of the simulated SMB is expected to be sensitive to grid resolution. Figure~\ref{fig:prect} shows the average grid spacing over the Greenland Ice Sheet (GrIS) in all six grids in this study. The $ne30pg2$ grid has the coarsest representation with an average $\Delta x=160~km$, and the $Arctic-GrIS$ grid has the highest resolution with an average $\Delta x=14.6~km$, similar to the grid spacing of the $11~km$ RACMO2.3 grid. The $ne30pg3$ grid has an average $\Delta x=111.2~km$, substantially coarser than the $f09$ grid, with an average $\Delta x=60~km$. Although $ne30pg3$ and $f09$ have similar average grid spacing over the entire globe, and comparable computational costs, the convergence of meridians on the FV grid enhances the resolution over the GrIS. The $Arctic$ grid has an average grid spacing of $\Delta x=27.8~km$, and is about 10 times more expensive than the $1^{\circ}$ models.  The $Arctic-GrIS$ grid is about twice as expensive as the $Arctic$ grid.

@@

The combined annual snow/ice melt shown in the bottom panel of Figure~\ref{fig:tseries} indicates that the $Arctic$ grid simulates the most realistic melt rates, with the other grids having more melt than RACMO. The $Arctic-GrIS$ grid overpredicts melting by about 125 Gt/yr. This is likely due to an anomalously warm lower troposphere during the summer, relative to the $Arctic$ run (Figure~\ref{fig:dThyps}). The $f19$ and $f09$ melting rates are improved over $Arctic-GrIS$, overestimating melt by only 70--90 Gt/yr. The SE grids have the largest positive melt bias, between 200--220 Gt/yr. It is more difficult to attribute these differences to resolution alone, since the FV grids have colder summer temperatures than the uniform-resolution SE grids.

@@

Figure~\ref{fig:basin} shows the basin-integrated PDD melt estimate.  In the large southeast and southwest basins (and all the other western basins), the $ne30pg3$ grid has larger PDD-based melt than the $ARCTCGRIS$ grid. The FV grids also have large PDD-based melt in the southwest basin, relative to $Arctic-GrIS$. The PDD plots indicate that the near-surface temperatures which contribute to melt are not well approximated by the summer lower-troposphere temperatures in Figure~\ref{fig:dThyps}.

@@

Figure~\ref{fig:ztrans} also shows the ice margin boundary, illustrating that the ablation zone lies in a narrow horizontal band where the ice sheet rapidly plunges to sea-level. Due to this abrupt transition, coarse grids will commonly represent the ablation zone with grid cells containing mixtures of ice-covered and ice-free regions. We hypothesize that coarser models have larger melt biases
%despite the $ARCTCIGRIS$ having the warmest troposphere,
because summer melting is confined to these mixed ice/land/ocean grid cells. CLM deals with land heterogeneity in a complex and sophisticated manner, but CAM only sees a homogenized state due to volume averaging over the sub-grid mixture. Thus, warm ice-free land patches in a grid cell may unduly influence the climate over the entire grid cell, causing a warm bias over the ice-covered patch.

Figure~\ref{fig:bias} shows mean melt bias, relative to both RACMO datasets, conditionally sampled based on grid cell ice fraction in the GrIS region. Errors are computed using the common-ice-mask approach, meaning that all fields are mapped to the common masks, which define the grid cell ice fraction. The figure shows that coarser grids generally have two peaks in ice fraction space; a bump in positive melting errors in the 0-20$\%$ range, and another in the fully ice-covered cells. Also shown are the $\pm 1$~standard deviation of the biases for each bin. They indicate that the biases in 0-20$\%$ bins are mostly contained in the positive bias region (a fractional bias greater than 0), whereas the fully-covered ice cells have a wider distribution, with many grid cells also containing negative melting biases. The excessive melting in the 0-20$\%$ ice fraction bins supports our hypothesis that the prevalence of mixed-grid cells in the ablation zone on coarse grids is responsible for their large melt bias.

@@

At the regional level, there is a halo of negative cloud bias around the ocean perimeter of Greenland on all $1-2^{\circ}$ grids, but not the VR grids. This halo bias coincides with a positive cloud bias over the ice sheet interior. This pattern has been traced to the inadequacy of the coarser grids for resolving orographic precipitation.  With overly smooth topography on the $1-2^{\circ}$ grids, synoptic systems moving into Greenland are not sufficiently lifted when encountering the steep ice margins.
As a result, moisture penetrates and dumps excess precipitation into the GrIS interior, instead of being concentrated near the coastal margins as shown by observations. This results in a positive precipitation and cloud bias in the ice sheet interior, and a halo of low cloud bias about the perimeter. The agreement of different observational data products on this bias lends confidence in the attribution of causes.
The VR grids compare better to the observations and show that orographic precipitation in Greenland is largely resolved.

>>>>

Before delving into the simulated Arctic, we assess the global mean differences between the various grids and dycores. Figure~\ref{fig:dT-lores} shows 1979-1998 annual mean, zonal mean height plots expressed as differences between uniform-resolution grids and dycores. The $f09$ grid is warmer than the $f19$ grid, primarily in the mid-to-high latitudes throughout the depth of the troposphere. This is a common response to increasing horizontal resolution in GCMs \cite{PS2002CD,RETAL2006JC}.
\citeA{HK2020QJRMS} have shown that this occurs in CAM due to greater resolved vertical velocities which, in turn, drive more condensational heating in the CLUBB macrophyiscs. The right columns in Figure~\ref{fig:dT-lores} support this interpretation, showing an increase in the climatological CLUBB heating in low ({\color{blue}{looks more like subtropics}}) and mid-latitudes on the $f09$ grid.

@@

It is useful to understand summer temperature biases due to their control on ice and snow melt over the Greenland Ice Sheet \cite<GrIS;>{O2001JAM}. Figure~\ref{fig:dThyps} shows the 1979--1998 lower troposphere summer temperature bias relative to ERA5. It is computed from the 500--1000 hPa geopotential thickness, solving for the layer-mean virtual temperature using the hypsometric equation ({\color{blue}{Might be good to write this down, for those of us who studied atmospheric science a really long time ago}}). The results are consistent with the zonal mean height plots; increasing resolution from $f19$ to $f09$ warms the climate, and the $1^{\circ}$ SE grids are warmer than the FV grids. The FV summer temperatures are persistently colder than ERA5, whereas the $1^{\circ}$ SE grids are not as cold, and are actually warmer than ERA5 at high-latitudes, north of $75^{\circ}$. All grids show a north-south gradient in bias over Greenland, with the summer temperature bias more positive for the northern part of the ice sheet. This pattern is also evident in the 2m summer temperature bias over Greenland (not shown).

@@

Some of these temperature differences may be related to different summer shortwave cloud forcing for the various grids and dycores. Figure~\ref{fig:SWCF} shows the summer shortwave cloud forcing bias in six runs, using the CERES-EBAF product. 
A negative bias corresponds to excessive reflection and cooling.
The uniform grids have similar biases, with the clouds reflecting 20--40 W/m$^2$ too much shortwave radiation over a wide swath of the Arctic, primarily the land masses ({\color{blue}{please explain what SWCF in caption.}}). There is also a halo of positive bias (clouds not reflective enough) around the ocean perimeter of Greenland. The $Arctic$ grid has much smaller cloud forcing biases over the Arctic land masses, but is still too reflective over Alaska, the Canadian Archipelago, and parts of Eurasia.  Compared to the $Arctic$ grid, the $ARCTCGRIS$ grid vastly reduces the cloud forcing bias over Eurasia, and also improves the bias over North America. In both VR grids, the halo of positive forcing bias around the perimeter of Greenland is absent.

While the summer cloud forcing biases are consistent with the summer temperature biases in Figure~\ref{fig:dThyps} -- regions where clouds are too reflective coincide with regions that are too cold -- it is not clear whether the cold biases are caused by the cloud biases, or whether the cold biases amplify the cloud forcing bias. ({\color{blue}{This ends abruptly. Anything else to add?}})

@@

The accuracy of the simulated SMB is expected to be sensitive to grid resolution. Figure~\ref{fig:prect}b shows the average grid spacing over the Greenland Ice Sheet (GrIS) in all six grids in this study. The $ne30pg2$ grid has the coarsest representation with an average $\Delta x=160~km$, and the $Arctic-GrIS$ grid has the highest resolution with an average $\Delta x=14.6~km$, similar to the grid spacing of the $11~km$ RACMO2.3 grid. The $ne30pg3$ grid has an average $\Delta x=111.2~km$, substantially coarser than the $f09$ grid, with an average $\Delta x=60~km$. Although $ne30pg3$ and $f09$ have similar average grid spacing over the entire globe, and comparable computational costs, the convergence of meridians on the FV grid enhances the resolution over the GrIS. The $Arctic$ grid has an average grid spacing of $\Delta x=27.8~km$, and is about 10 times more expensive than the $1^{\circ}$ models.  The $Arctic-GrIS$ grid is about twice as expensive as the $Arctic$ grid. ({\color{blue}{Wondering if this paragraph would go better in an earlier section.}})

@@

The combined annual snow/ice melt shown in the bottom panel of Figure~\ref{fig:tseries} indicates that the $Arctic$ grid simulates the most realistic melt rates, with the other grids having more melt than RACMO. The $Arctic-GrIS$ grid overpredicts melting by about 125 Gt/yr. This is likely due to an anomalously warm lower troposphere during the summer, relative to the $Arctic$ run (Figure~\ref{fig:dThyps}). The $f19$ and $f09$ melting rates are improved over $Arctic-GrIS$, overestimating melt by only 70--90 Gt/yr. The SE grids have the largest positive melt bias, between 200--220 Gt/yr. It is more difficult to attribute these differences to resolution alone, since the FV grids have colder summer temperatures than the uniform-resolution SE grids. ({\color{blue}{Not sure I understand this sentence.  FV is both cooler and higher-resolution than SE, so one might suspect that cooler T is, in fact, connected to higher resolution.}})

@@

Figure~\ref{fig:basin}c shows the basin-integrated PDD melt estimate.  In the large southeast and southwest basins (and all the other western basins), the $ne30pg3$ grid has larger PDD-based melt than the $ARCTCGRIS$ grid. The FV grids also have large PDD-based melt in the southwest basin, relative to $Arctic-GrIS$. The PDD plots indicate that the near-surface temperatures which contribute to melt are not well approximated by the summer lower-troposphere temperatures in Figure~\ref{fig:dThyps}.

@@

Figure~\ref{fig:ztrans} also shows the ice margin boundary, illustrating that the ablation zone lies in a narrow horizontal band where the ice sheet rapidly plunges to sea-level. Due to this abrupt transition, coarse grids will commonly represent the ablation zone with grid cells containing mixtures of ice-covered and ice-free regions. We hypothesize that coarser models have larger melt biases
%despite the $ARCTCIGRIS$ having the warmest troposphere,
because summer melting is confined to these mixed ice/land/ocean grid cells. CLM deals with land heterogeneity in a complex and sophisticated manner, but CAM only sees a homogenized state due to volume averaging over the sub-grid mixture. Thus, warm ice-free land patches in a grid cell may unduly influence the climate over the entire grid cell, causing a warm bias over the ice-covered patch. ({\color{blue}{This is an interesting conclusion pointing to the need for better treatment of surface inhomogeneity in CAM.  This might be a way to compensate for coarse resolution in future CESM versions?}})

Figure~\ref{fig:bias} shows mean melt bias, relative to both RACMO datasets, conditionally sampled based on grid cell ice fraction in the GrIS region. Errors are computed using the common-ice-mask approach, meaning that all fields are mapped to the common masks, which define the grid cell ice fraction. The figure shows ({\color{blue}{Any idea why the errors are smaller in the cells with intermediate ice fraction?  I wouldn't have expected this.}}) that coarser grids generally have two peaks in ice fraction space; a bump in positive melting errors in the 0-20$\%$ range, and another in the fully ice-covered cells. Also shown are the $\pm 1$~standard deviation of the biases for each bin. They indicate that the biases in 0-20$\%$ bins are mostly contained in the positive bias region (a fractional bias greater than 0), whereas the fully-covered ice cells have a wider distribution, with many grid cells also containing negative melting biases. The excessive melting in the 0-20$\%$ ice fraction bins supports our hypothesis that the prevalence of mixed-grid cells in the ablation zone on coarse grids is responsible for their large melt bias.

@@

At the regional level, there is a halo of negative cloud bias (/color{blue}{I got confused about signs here, because section 3 talks about a halo of positive cloud SW forcing bias, which corresponds to a negative bias in cloud amounts.  Maybe replace 'cloud bias' with 'cloudiness bias' or something similar?}) around the ocean perimeter of Greenland on all $1-2^{\circ}$ grids, but not the VR grids. This halo bias coincides with a positive cloud bias over the ice sheet interior. This pattern has been traced to the inadequacy of the coarser grids for resolving orographic precipitation.  With overly smooth topography on the $1-2^{\circ}$ grids, synoptic systems moving into Greenland are not sufficiently lifted when encountering the steep ice margins.
As a result, moisture penetrates and dumps excess precipitation into the GrIS interior, instead of being concentrated near the coastal margins as shown by observations. This results in a positive precipitation and cloud bias in the ice sheet interior, and a halo of low cloud bias about the perimeter. The agreement of different observational data products on this bias lends confidence in the attribution of causes.
The VR grids compare better to the observations and show that orographic precipitation in Greenland is largely resolved.
